# SECU0057 Data Science Project: Text Mining and Machine Learning

**Candidate Number:** RWFP0

This R Markdown notebook reproduces the text mining and machine learning part of the project.

**Input**

- `data/processed/combined_labeled.csv`

**Output**

- `outputs/tables/svm_confusion_matrix.csv`
- `outputs/tables/svm_metrics.csv`
- `outputs/tables/top_tfidf_terms_by_label.csv`
- `outputs/tables/top_bigram_tfidf_terms_by_label.csv`
- `outputs/figures/*.png`

**Notes**

- A fixed random seed is used for reproducibility.
- The vocabulary is built only from training data to avoid data leakage.
- All paths are relative to the project root folder.

# 1. Load packages

```{r}
# !!Install packages only if they are missing
# install.packages(c("dplyr", "stringr", "tidytext", "ggplot2", "tidyr", "e1071"), repos = "https://cloud.r-project.org")

library(dplyr)
library(stringr)
library(tidytext)
library(ggplot2)
library(tidyr)
library(e1071)
```

# 2. Load labelled data

```{r}
# Load the processed dataset created by data_preprocessing_and_weak_labels.R.
data <- read.csv("data/processed/combined_labeled.csv")

cat("Rows:", nrow(data), "\n")

cat("Source distribution:\n")
print(table(data$source))


cat("Weak label distribution:\n")
print(table(data$weak_label))
```

# 3. Basic cleaning

```{r}
# Remove missing values, create document IDs, and standardise text fields.
data <- data |>
  filter(!is.na(text), !is.na(weak_label)) |>
  mutate(
    doc_id = row_number(),
    weak_label = as.factor(weak_label),
    text = str_squish(text),
    text_lower = tolower(text)
  )


cat("Rows after cleaning:", nrow(data), "\n")
print(table(data$weak_label))
```

# 4. Train and test split

```{r}
# Use a stratified 80/20 split so each weak label appears in both sets.
set.seed(5757) # For reproducibility


data <- data |>
  mutate(row_id = row_number())

train_ids <- data |>
  group_by(weak_label) |>
  sample_frac(0.8) |>
  pull(row_id)

train_data <- data |> filter(row_id %in% train_ids)
test_data <- data |> filter(!row_id %in% train_ids)

cat("Train label distribution:\n")
print(table(train_data$weak_label))

cat("Test label distribution:\n")
print(table(test_data$weak_label))
```

# 5. Create TF-IDF vocabulary

```{r}
# Tokenise the training text and remove stopwords.
# The vocabulary is built only from the training data to avoid data leakage.
tokens_train <- train_data |>
  select(doc_id, text) |>
  unnest_tokens(word, text) |>
  filter(
    str_detect(word, "^[a-z]+$"),
    nchar(word) > 2
  ) |>
  anti_join(stop_words, by = "word")

# Calculate TF-IDF values in the training set.
tfidf_train <- tokens_train |>
  count(doc_id, word, sort = FALSE) |>
  bind_tf_idf(word, doc_id, n)

# Keep the top 500 terms as the model vocabulary.
top_terms <- tfidf_train |>
  group_by(word) |>
  summarise(total_tfidf = sum(tf_idf), .groups = "drop") |>
  arrange(desc(total_tfidf)) |>
  slice_head(n = 500)

vocab <- top_terms$word

cat("Vocabulary size:", length(vocab), "\n")
```

# 6. Create TF-IDF matrices
# This function creates a TF-IDF matrix using the fixed training vocabulary.
# It adds missing vocabulary columns so train and test matrices have the same columns.
# This avoids the "test data does not match model" error.

```{r}
# It adds missing vocabulary columns so train and test matrices match.
make_tfidf_matrix <- function(input_data, vocab_terms) {


  token_data <- input_data |>
    select(doc_id, text) |>
    unnest_tokens(word, text) |>
    filter(
      str_detect(word, "^[a-z]+$"),
      nchar(word) > 2,
      word %in% vocab_terms
    ) |>
    anti_join(stop_words, by = "word")
  
  tfidf_data <- token_data |>
    count(doc_id, word, sort = FALSE) |>
    bind_tf_idf(word, doc_id, n)
  
  wide_data <- tfidf_data |>
    select(doc_id, word, tf_idf) |>
    pivot_wider(
      names_from = word,
      values_from = tf_idf,
      values_fill = 0
    )
  
  output <- input_data |>
    select(doc_id, weak_label) |>
    left_join(wide_data, by = "doc_id")
  
  output[is.na(output)] <- 0
  
  missing_terms <- setdiff(vocab_terms, names(output))
  
  for (term in missing_terms) {
    output[[term]] <- 0
  }
  
  output <- output |>
    select(doc_id, weak_label, all_of(vocab_terms))
  
  return(output)
}

train_matrix <- make_tfidf_matrix(train_data, vocab)
test_matrix <- make_tfidf_matrix(test_data, vocab)

x_train <- train_matrix |> select(-doc_id, -weak_label)
y_train <- train_matrix$weak_label

x_test <- test_matrix |> select(-doc_id, -weak_label)
y_test <- test_matrix$weak_label

cat("Training matrix dimensions:\n")
print(dim(x_train))

cat("Testing matrix dimensions:\n")
print(dim(x_test))
```

# 7. Train linear SVM model

```{r}
# Train a linear SVM because it works well with sparse text features
# and is easier to explain than a black-box model.
svm_model <- svm(
  x = x_train,
  y = y_train,
  kernel = "linear"
)

pred <- predict(svm_model, x_test)
```

# 8. Evaluate model

```{r}
# Create confusion matrix.
conf_mat <- table(
  Predicted = pred,
  Actual = y_test
)

print(conf_mat)

# Calculate accuracy.
accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
cat("Accuracy:", accuracy, "\n")

# Calculate macro-F1 manually.
# Macro-F1 is useful because the weak labels are not perfectly balanced.
calculate_f1 <- function(conf_mat) {
  classes <- colnames(conf_mat)
  f1_scores <- c()
  
  for (class in classes) {
    tp <- conf_mat[class, class]
    fp <- sum(conf_mat[class, ]) - tp
    fn <- sum(conf_mat[, class]) - tp
    
    precision <- ifelse((tp + fp) == 0, 0, tp / (tp + fp))
    recall <- ifelse((tp + fn) == 0, 0, tp / (tp + fn))
    
    f1 <- ifelse(
      (precision + recall) == 0,
      0,
      2 * precision * recall / (precision + recall)
    )
    
    f1_scores <- c(f1_scores, f1)
  }
  
  return(mean(f1_scores))
}

macro_f1 <- calculate_f1(conf_mat)
cat("Macro-F1:", macro_f1, "\n")




```

# 9. Save output tables

```{r}
# Create output folders.
dir.create("outputs", showWarnings = FALSE)
dir.create("outputs/tables", showWarnings = FALSE)
dir.create("outputs/figures", showWarnings = FALSE)

# Save confusion matrix.
conf_df <- as.data.frame(conf_mat)

write.csv(
  conf_df,
  "outputs/tables/svm_confusion_matrix.csv",
  row.names = FALSE
)

# Save model metrics.
metrics_df <- data.frame(
  model = "linear_svm_tfidf",
  accuracy = accuracy,
  macro_f1 = macro_f1,
  vocabulary_size = length(vocab),
  train_rows = nrow(train_data),
  test_rows = nrow(test_data)
)



write.csv(
  metrics_df,
  "outputs/tables/svm_metrics.csv",
  row.names = FALSE
)

print(metrics_df)
```

# 10. Weak label distribution plots

```{r}
# Count plot: useful for showing raw class counts by source.
label_count_plot <- data |>
  count(source, weak_label) |>
  ggplot(aes(x = weak_label, y = n, fill = source)) +
  geom_col(position = "dodge") +
  labs(
    title = "Distribution of weak stress-narrative labels by source",
    x = "Weak label",
    y = "Number of texts"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

label_count_plot

ggsave(
  "outputs/figures/weak_label_distribution.png",
  label_count_plot,
  width = 8,
  height = 5
)

# Proportion plot: better for comparing Guardian and Ethereum Research
# because the two sources have different sample sizes.
label_prop_plot <- data |>
  count(source, weak_label) |>
  group_by(source) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = weak_label, y = prop, fill = source)) +
  geom_col(position = "dodge") +
  labs(
    title = "Proportion of weak stress-narrative labels by source",
    x = "Weak label",
    y = "Proportion within source"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

label_prop_plot

ggsave(
  "outputs/figures/weak_label_distribution_proportion.png",
  label_prop_plot,
  width = 8,
  height = 5
)
```

# 11. Top TF-IDF terms by weak label

```{r}
# These custom stopwords remove some names and technical tokens
# that were not useful for interpreting the stress narratives.
custom_stopwords <- c(
  "trump", "farage", "president", "minister", "ministers",
  "liberal", "nsw", "police",
  "fei", "fhe", "rai", "solo", "pid",
  "gemini", "metacoin", "ghostpool",
  "slice", "heap", "ssusd",
  "preconfer", "preconf", "preconfirmations", "preconfers",
  "claimable", "forkable", "rebased",
  "stamina"
)

label_tokens <- data |>
  select(doc_id, weak_label, text) |>
  unnest_tokens(word, text) |>
  filter(
    str_detect(word, "^[a-z]+$"),
    nchar(word) > 2,
    !word %in% custom_stopwords
  ) |>
  anti_join(stop_words, by = "word")

label_tfidf <- label_tokens |>
  count(weak_label, word, sort = FALSE) |>
  bind_tf_idf(word, weak_label, n)

top_label_terms <- label_tfidf |>
  group_by(weak_label) |>
  arrange(desc(tf_idf)) |>
  slice_head(n = 10) |>
  ungroup()

print(top_label_terms)

write.csv(
  top_label_terms,
  "outputs/tables/top_tfidf_terms_by_label.csv",
  row.names = FALSE
)

top_terms_plot <- top_label_terms |>
  mutate(word = reorder_within(word, tf_idf, weak_label)) |>
  ggplot(aes(x = word, y = tf_idf, fill = weak_label)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ weak_label, scales = "free") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    title = "Top TF-IDF terms by weak stress-narrative label",
    x = "Term",
    y = "TF-IDF"
  ) +
  theme_minimal()

top_terms_plot

ggsave(
  "outputs/figures/top_tfidf_terms_by_label.png",
  top_terms_plot,
  width = 10,
  height = 7
)
```

# 12. Bigram TF-IDF analysis

```{r}
# Bigram analysis is used because two-word phrases are often easier
# to interpret than single words.
custom_stopwords_bigram <- c(
  "trump", "farage", "president", "minister", "ministers",
  "liberal", "nsw", "fei", "fhe", "rai", "solo", "pid",
  "gemini", "metacoin", "ghostpool",
  "slice", "heap", "ssusd",
  "preconfer", "preconf", "preconfirmation",
  "preconfirmations", "preconfers",
  "claimable", "forkable", "rebased",
  "stamina", "justindrake", "laplace", "gamma", "asic",
  "skde", "msgs", "pubkeys"
)

bigram_tokens <- data |>
  select(doc_id, weak_label, text) |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate(bigram, into = c("word1", "word2"), sep = " ") |>
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word,
    !word1 %in% custom_stopwords_bigram,
    !word2 %in% custom_stopwords_bigram,
    str_detect(word1, "^[a-z]+$"),
    str_detect(word2, "^[a-z]+$"),
    nchar(word1) > 2,
    nchar(word2) > 2
  ) |>
  unite(bigram, word1, word2, sep = " ")

bigram_tfidf <- bigram_tokens |>
  count(weak_label, bigram, sort = FALSE) |>
  bind_tf_idf(bigram, weak_label, n)

top_bigram_terms <- bigram_tfidf |>
  group_by(weak_label) |>
  arrange(desc(tf_idf)) |>
  slice_head(n = 8) |>
  ungroup()

print(top_bigram_terms)

write.csv(
  top_bigram_terms,
  "outputs/tables/top_bigram_tfidf_terms_by_label.csv",
  row.names = FALSE
)

bigram_plot <- top_bigram_terms |>
  mutate(bigram = reorder_within(bigram, tf_idf, weak_label)) |>
  ggplot(aes(x = bigram, y = tf_idf, fill = weak_label)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ weak_label, scales = "free") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    title = "Top bigram TF-IDF terms by weak stress-narrative label",
    x = "Bigram",
    y = "TF-IDF"
  ) +
  theme_minimal()

bigram_plot

ggsave(
  "outputs/figures/top_bigram_tfidf_terms_by_label.png",
  bigram_plot,
  width = 10,
  height = 7
)
```

# 13. Confusion matrix heatmap

```{r}
# This figure visualises where the SVM classifier is correct
# and where weak labels are confused.
label_order <- c(
  "fraud_security",
  "market_panic",
  "neutral_other",
  "regulatory_stress",
  "technical_friction"
)

conf_df <- read.csv("outputs/tables/svm_confusion_matrix.csv")

if (!all(c("Predicted", "Actual", "Freq") %in% names(conf_df))) {
  colnames(conf_df)[1:3] <- c("Predicted", "Actual", "Freq")
}

conf_df <- conf_df |>
  mutate(
    Predicted = factor(Predicted, levels = label_order),
    Actual = factor(Actual, levels = label_order)
  )

conf_plot <- ggplot(conf_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 3.2) +
  scale_y_discrete(limits = rev(label_order)) +
  labs(
    title = "SVM confusion matrix",
    x = "Predicted label",
    y = "Actual label"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 11),
    legend.position = "none"
  )

conf_plot

ggsave(
  "outputs/figures/confusion_matrix_heatmap.png",
  conf_plot,
  width = 6.5,
  height = 5
)
```

# 14. Baseline comparison plot

```{r}
# This plot compares the SVM accuracy with simple baselines.
# It helps show that the model performs above simple guessing.
# Calculate simple baselines from the test set.
random_baseline <- 1 / length(unique(y_test))
largest_class_baseline <- max(table(y_test)) / length(y_test)

performance_df <- data.frame(
  model = c("Random baseline", "Largest-class baseline", "Linear SVM"),
  accuracy = c(random_baseline, largest_class_baseline, accuracy)
)
performance_plot <- ggplot(performance_df, aes(x = model, y = accuracy)) +
  geom_col() +
  geom_text(aes(label = round(accuracy, 3)), vjust = -0.4, size = 4) +
  ylim(0, 0.7) +
  labs(
    title = "Model accuracy compared with simple baselines",
    x = "Model",
    y = "Accuracy"
  ) +
  theme_minimal()

performance_plot

ggsave(
  "outputs/figures/model_accuracy_baseline_comparison.png",
  performance_plot,
  width = 7,
  height = 4
)
```

# 15. Text length distribution by source
```{r}
# This section checks whether Guardian and Ethereum Research texts
# have different text lengths. This supports the limitation that
# the two sources are different types of text.

text_length_summary <- data |>
  mutate(text_length = nchar(text)) |>
  group_by(source) |>
  summarise(
    n = n(),
    mean_length = mean(text_length),
    median_length = median(text_length),
    min_length = min(text_length),
    max_length = max(text_length),
    .groups = "drop"
  )

print(text_length_summary)

write.csv(
  text_length_summary,
  "outputs/tables/text_length_summary_by_source.csv",
  row.names = FALSE
)

text_length_plot <- data |>
  mutate(text_length = nchar(text)) |>
  ggplot(aes(x = source, y = text_length, fill = source)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Text length distribution by source",
    x = "Source",
    y = "Number of characters"
  ) +
  theme_minimal()

text_length_plot

ggsave(
  "outputs/figures/text_length_distribution_by_source.png",
  text_length_plot,
  width = 6,
  height = 4
)

```
# The end of the notebook. All outputs are saved in the outputs folder.